```{r setup, include=FALSE}
# set up code options for Rmd 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

# Contemporary environmental justice and wildlife conservation issue in Los Angeles County, California 
## Background

Contemporary environmental justice often reflects historical injustices, especially in the persistent legacy of racial segregation in the United States. During the 1930s, the Home Owners' Loan Corporation (HOLC) evaluated neighborhoods under the New Deal based on perceived real estate investment safety, using a discriminatory "redlining" system of A (green), B (blue), C (yellow), and D (red) to limit home loan access. This is covered by the [New York Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html). This practice, extensively documented, has not only impacted community wealth but also health by creating areas with fewer green spaces and higher temperatures. Recent research has shown that redlining not only affects local environments but also skews biodiversity observations. Citizen science, a platform for sharing species observations, generates substantial data, yet Ellis-Soto and colleagues discovered that redlined neighborhoods are significantly underrepresented across 195 US cities in these observations. This disparity is concerning as conservation decisions heavily rely on such data for effective policies and actions. This is covered by [EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

## Goal

Understanding these connections can aid in developing targeted interventions or policies to address both environmental justice issues and wildlife conservation in these areas.

## Datsets

```{r load-packages}
library(tidyverse)
library(sf)
library(tmap)
rm(list = ls())
```

#### EJScreen

The US EPA provides EJScreen as a tool for communities, stakeholders, and researchers to access environmental and demographic information, aiming to foster transparency, inform decisions on pursuing environmental justice, and establish a shared platform between the agency and the public for addressing environmental justice concerns, and filter to LA county.

Census track: Census Tracts are essential for collecting and presenting statistical data in various surveys, including the Census, to analyze trends and make informed policy decisions

Block Group levels: Block Groups are subdivisions within Census Tracts, offering a more detailed breakdown of population characteristics.

```{r load-data}
# read in EJScreen data
ejscreen <- st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/") 

# filter to LA county
LA_ejscreen <- ejscreen %>%
  filter(CNTY_NAME %in% c("Los Angeles County"))
```

#### Mapping Inequality

Under the Mapping Inequality project, researchers, headed by the Digital Scholarship Lab at the University of Richmond, have digitized HOLC maps and related information. Specifically, we will be focusing on working with HOLC grade designations maps for Los Angeles. Further details about the data can be accessed here.

```{r import-redlining}
# read the HOLC map 
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>% 
  st_make_valid()
```

#### Biodiversity observations

The Global Biodiversity Information Facility (GBIF) stands as the world's largest repository for biodiversity observations, encompassing data such as species sightings along with their respective locations and dates, and filtering 2022 dataset.

```{r load-bird-data}
# read in bird data
gbif <- st_read("data/gbif-birds-LA/") %>%
  filter(year == 2022) # filter to 2022

# transform data to match CRS
gbif <- st_transform(gbif, crs = st_crs(LA_redlining))
```

#### 

## Analysis

### Persistent Impact of Redlining on Present-day Environmental (In)justice

Displaying wastewater discharge across census block groups. Identify the census block groups that exceed the 95th percentile of wastewater discharge by marking their centroids.

```{r plot-wastewater}
# create centroids of LA EJScreen data
# filter to census block groups > 95th wastewater discharge
LA_ejscreen_centroids <- st_centroid(LA_ejscreen) %>%
  filter(P_PWDIS > 95)

# make map of wastewater discharge
map1 <- tm_shape(LA_ejscreen) +
  tm_fill(fill = "PWDIS", title = "Wastewater discharge") +
  tm_graticules()

# add centroids to map
map1 +
  tm_shape(LA_ejscreen_centroids) +
  tm_dots()
```

**Vulnerable People**: Classify the population considered low income as less than 5%.

```{r filter2}
# Filtering the LA_ejscreen dataset for areas with high Particulate Matter 2.5 (P_PM25) and high Percent of Population Below Poverty Level (P_PNPL)
LA_ejscreen_inequal <- LA_ejscreen %>% 
  filter(P_PM25 > 80 & P_PNPL > 80) %>%
  nrow() / nrow(LA_ejscreen) * 100

# Displaying the percentage of the population impacted by environmental quality in Los Angeles
print(paste("Population considered low income as less than 5% in LA:",LA_ejscreen_low))
```

Investigates areas with high levels of both Particulate Matter 2.5 and Superfund sites, examining how their coexistence affects the environment in specific regions.

```{r filter1}
# filter pm2.5 80% and superfund site
LA_ejscreen_inequal <- LA_ejscreen %>% 
  filter(P_PM25 > 80) %>%
  filter(P_PNPL > 80) %>%
  nrow()/nrow(LA_ejscreen)*100

print(paste("Population impacted by environmental quality in LA:", LA_ejscreen_inequal))
```

### Map delineating historical redlining boundaries and color-code them based on HOLC grades

```{r map-redlining}
# map with color grade 
tm_shape(LA_redlining) +
  tm_fill("grade") +
  tm_graticules() 
```

The number of census block groups situated within HOLC-graded areas while ensuring the alignment of coordinate reference systems (CRS). The original prompt was open to different topological relationships, resulting in varied interpretations. To illustrate, utilizing "intersects" identifies all census block groups that overlap with redlined areas, totaling 3951 CBGs. Alternatively, employing "within" exclusively selects census block groups contained within redlined areas, amounting to 887 CBGs

```{r ej-red-transform}
# first transform data to same CRS
LA_ejscreen <- st_transform(LA_ejscreen, crs= st_crs(LA_redlining))

# map both datasets to get a sense of their spatial pattern
plot(LA_ejscreen["ID"])
plot(LA_redlining["grade"])
```

#### Number of census block group

```{r ej-red2}
# try using the st_intersects functions
sel_sgbp <- st_intersects(x = LA_ejscreen, y = LA_redlining) # returns binary predicate list
sel_logical = lengths(sel_sgbp) > 0 # create logical of which items do intersect
redlining_cbgs2 <- LA_ejscreen[sel_logical, ] # filter census block groups based on logicals

# the number of census block groups
print(paste("the number of census block groups:", nrow(redlining_cbgs2)))

```

### Table

present conditions within historical redlining categories using the mean values of the following variables: percent low income, percentile for Particulate Matter 2.5, percentile for low life expectancy, and percentile for air toxics cancer risk based on EJScreen data.

```{r ej-red-summary}
# summarize info with use st_intersection 
LA_ejscreen <- st_transform(LA_ejscreen, crs= st_crs(LA_redlining))

LA <- st_intersection(LA_redlining, LA_ejscreen)

LA %>%
  group_by(grade) %>%
  summarise(lowincpct = mean(LOWINCPCT, na.rm = TRUE),
            pm25 = mean(P_PM25, na.rm = TRUE),
            lifeexppct = mean(P_LIFEEXPPCT, na.rm = TRUE),
            cancer = mean(P_CANCER, na.rm = TRUE))
```

### Impact of redlining on biodiversity observations

For bird observations recorded in 2022 within neighborhoods categorized by HOLC grades, calculate the percentage of observations within each redlining category, and visualize the outcomes.

```{r gbif-red1}
# try st_join with st_intersects
gbif1 <- st_join(x = gbif, y = LA_redlining, join = st_intersects, left = FALSE)

gbif_summary1 <- gbif1 %>% 
  st_set_geometry(NULL) %>%
  group_by(grade) %>%                  
  summarise(count = n()) %>%      
  mutate(percentage = (count / sum(count))*100 )

ggplot(data = gbif_summary1) +
  geom_bar(aes(x = grade, y = percentage), stat = "identity") +
  labs(x = "HOLC grade", y = "Percentage of observations", 
       title = "Percentage of Bird Observations in Redlining Categories") 
```

## Conclusion

The analysis of these three datasets unveils a striking association between vulnerable communities and bird populations. There is a clear trend indicating that vulnerable populations coincide with specific patterns in bird populations.

It becomes apparent that areas graded as 'C and D' in the HOLC system exhibit higher vulnerability compared to other communities. These 'C and D' HOLC grade regions consistently harbor significantly larger low-income populations, more than doubling the figures seen in 'A' graded areas. Additionally, these same areas face elevated levels of air pollution, lower life expectancies, and an increased incidence of cancer compared to neighborhoods with different HOLC grades.

The intriguing aspect lies in the observation that a notable number of birds are sighted in HOLC grade 'D' areas, despite the prevalent high levels of air pollution. It is plausible to speculate that 'C' and 'D' graded areas may possess abundant food waste sources in their streets, potentially explaining the heightened bird observations in these neighborhoods. This correlation between vulnerability metrics and bird populations raises interesting questions regarding environmental dynamics and resource availability within these communities.
